---
title: "Random effects"
author: "Ashlee Mikkelsen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}

rm(list = ls())
setwd("~/Rprojects/Diet-and-reproduction-in-rown-bears")

library(ggplot2)
library(dplyr)
library(GGally)
library(viridis)
library(INLA)
library(brinla)
library(usdm)
library(wiqid)

P <- palette(viridis(8))
bear.reprodata <- read.csv("ReproData_20220503.csv")
bear.reprodata$fYEAR <- as.factor(bear.reprodata$fYEAR)
bear.reprodata$Cmonth <- as.factor(bear.reprodata$Cmonth)
str(bear.reprodata)

```


## Hypothesis #1: Explaining variation in length of maternal care

Maternal care = b0+ (b1 x cortisol)+(b2 x d15N)+(b3 x body condition)+(b4 x litter size) + (b5 x offspring condition) | BearID + family group

I move through this analysis following Zuur et al. 2013.

### Data exploration

#### categorical variables
```{r}
#Number of observations per maternal care
table(bear.reprodata$length.mat.care)
# 48 had cubs die, 12 females died causing unnatural seperation, 98 need to be removed because they refer to years when there was no reproduction, 254 cared for litters for 1.5 years and 123 cared for litters for 2.5 years.

hyp1.dat <- bear.reprodata[which(bear.reprodata$length.mat.care!="n"),]
hyp1.dat <- hyp1.dat[which(hyp1.dat$length.mat.care!="cd"),]
hyp1.dat <- hyp1.dat[which(hyp1.dat$length.mat.care!="d"),]

# Number of observations per litter size
table(hyp1.dat$litter.size)

# There is one record in which there was an unknown litter size (0), 46 singles, 184 twins, 137 triplets, and 9 quadruplets. The small quadruplet size is likely to give us some issues and may need to be combined with triplets, but we will leave them separate first.

hyp1.dat <- hyp1.dat[which(hyp1.dat$litter.size>0),]

```

#### Continuous variables

```{r}
hyp1.dat <- subset(hyp1.dat, repro=="WithCubs")
#Oof. That really knocked down the sample size

# length of maternal care (cont)
dotchart(hyp1.dat$n.LMC)

# Cortisol
dotchart(hyp1.dat$cortisol.kitcorrected)
# We have 1 outlier with greater than 25. We leave it for now but need to be mindful of it

# d15N
dotchart(hyp1.dat$dN15)

#body condition
dotchart(hyp1.dat$condition)
# We have a bit of an extreme bi-modal distribution because this dataset includes females and offspring. I want to just focus on the females. *Unless cubs are part of the family seperation?*

# offspring condition
dotchart(hyp1.dat$mean.offspring.condition)

####### Any linear relationship?
ggplot(data = hyp1.dat,aes(condition, mean.offspring.condition))+
  geom_point()+
  geom_smooth()+
  theme_classic()
# unsurprisingly, we have a linear relationship between these two.
lmCondition <- lm(mean.offspring.condition~condition, data = hyp1.dat)
summary(lmCondition)
#r squared = 0.46. Not great, but not the end of the world.

ggplot(data = hyp1.dat, aes(condition, cortisol.kitcorrected))+
  geom_point()+
  theme_classic()+
  geom_smooth()
# we have another linear relationship
lmcort.cond <- lm(condition~cortisol.kitcorrected, data = hyp1.dat)
summary(lmcort.cond)
# r squared = 0.18. very interesting but fine t run togther in same model

ggplot(data = hyp1.dat, aes(dN15, mean.offspring.condition))+
  geom_point()+
  geom_smooth(method = "lm")+
  theme_classic()
# very slight linear relationship. Man, we have a lot of correlated variables.
lmN15.offspring <- lm(mean.offspring.condition~dN15, data = hyp1.dat)
summary(lmN15.offspring)
# It is nothing!
```


#### Colinearity
```{r}
myvar <-  c("age","dN15","dC13","cortisol.kitcorrected",
            "condition","n.LMC","litter.size","mean.offspring.condition")

Mypairs <- (hyp1.dat[,myvar])
vif(Mypairs)
#All of my VIFs are below 3, so I do not need to worry too much about covariance


############ End of data exploration

```


### Linear models

Prior to analysis, I standardize all continuous covariates

```{r}

data.hyp1 <- as.data.frame(
  cbind(
  "bearID"=hyp1.dat$bearID,
  "fYR"= hyp1.dat$fYEAR,
  "year"=as.numeric(hyp1.dat$year.rep),
  "Zage"=as.numeric(standardize(hyp1.dat$age)),
  "repro"=hyp1.dat$repro,
  "Zd15N"=as.numeric(standardize(hyp1.dat$dN15)),
  "Zd13C"=as.numeric(standardize(hyp1.dat$dC13)),
  "Zcort"=as.numeric(standardize(hyp1.dat$cortisol.kitcorrected)),
  "Zcondition"=as.numeric(standardize(hyp1.dat$condition)),
  "fg"=hyp1.dat$family.group,
  "fLMC"=hyp1.dat$length.mat.care,
  "Zlitter"=as.numeric(standardize(hyp1.dat$litter.size)),
  "Zmoc"=as.numeric(standardize(hyp1.dat$mean.offspring.condition))
  )
)

data.hyp1$year <- as.numeric(data.hyp1$year)
data.hyp1$Zage <- as.numeric(data.hyp1$Zage)
data.hyp1$Zd15N <- as.numeric(data.hyp1$Zd15N)
data.hyp1$Zd13C <- as.numeric(data.hyp1$Zd13C)
data.hyp1$Zcort <- as.numeric(data.hyp1$Zcort)
data.hyp1$Zcondition <- as.numeric(data.hyp1$Zcondition)
data.hyp1$Zlitter <- as.numeric(data.hyp1$Zlitter)
data.hyp1$Zmoc <- as.numeric(data.hyp1$Zmoc)

```


#### Univariate models

I was getting quite lost in this analysis yesterday and after a chat with Andreas, I still feel a bit lost, but following his advice I am going to start simple and just think about some fundamental questions. I don't need to build one giant model with every possible explanantion and account for every bit of variation.
```{r}

#####female Age
hyp1.femage <- inla(
  Zage~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.femage)

# There is evidence that length of maternal care varies by female age, where older females tend to spend more time with offspring
ggplot(data = data.hyp1, aes(fLMC,Zage))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized female bear age")+
  xlab("Length of maternal care")


##### Nitrogen

hyp1.nitro <- inla(
  Zd15N~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.nitro)
# no evidence that Nitrogen varies between maternal strategies
ggplot(data = data.hyp1, aes(fLMC,Zd15N))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized d 15N")+
  xlab("Length of maternal care")

##### Carbon

hyp1.carbon <- inla(
  Zd13C~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.carbon)
# No evidence that carbon varies by length of maternal care
ggplot(data = data.hyp1, aes(fLMC,Zd13C))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized d 13C")+
  xlab("Length of maternal care")

##### cortisol

hyp1.cort <- inla(
  Zcort~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.cort)
# very weak evidence that cortisol is higher for females that give 2 years of maternal care
ggplot(data = data.hyp1, aes(fLMC,Zcort))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized hair cortisol concentration")+
  xlab("Length of maternal care")


##### female body condition

hyp1.condition <- inla(
  Zcondition~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.condition)
# No evidence that female condition varies by length of maternal care
ggplot(data = data.hyp1, aes(fLMC,Zcondition))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized female condition")+
  xlab("Length of maternal care")


##### litter size

hyp1.litter <- inla(
  Zlitter~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.litter)
# No evidence that litter size varies by length of maternal care
ggplot(data = data.hyp1, aes(fLMC,Zlitter))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized litter size")+
  xlab("Length of maternal care")


##### mean offspring condition

hyp1.offspringcond <- inla(
  Zmoc~fLMC+
    f(bearID, model = "iid"),
  control.compute = list(dic=TRUE, waic=TRUE),
  control.predictor = list(compute=TRUE),
  quantiles = c(0.025, 0.975),
  data = data.hyp1,
  
)
summary(hyp1.offspringcond)
# No evidence that offspring condition varies by length of maternal care
ggplot(data = data.hyp1, aes(fLMC,Zmoc))+
  geom_boxplot()+
  theme_classic()+
  ylab("Standardized mean offspring condition")+
  xlab("Length of maternal care")


```

Rethinking my approach. Maybe the best way to go about this is to look at females that have been resampled

```{r}

resamples <- data.hyp1 %>%
  group_by(bearID) %>%
  summarise(
    n=n()
  )

RSfems <- subset(resamples,n>1)

Test.data <- subset(data.hyp1, bearID %in% RSfems$bearID)


ggplot(data = Test.data, aes(bearID, fLMC))+
  geom_jitter(height = 0.03, width = 0.03, size=2, alpha=0.5, aes(color=Zmoc))+
  scale_color_viridis()

```
We have this subset of 40 females that we re-sampled over multiple years from 2-7 times each. 21 females never stayed with cubs 2.5 years. 6 females always stayed with cubs 2.5 years. 14 females swapped between the two strategies. Why? Can diet estimates tell us something?


UGH! This length of maternal care business is tricky!


```{r}

ggplot(data = Test.data,aes(year, Zd15N))+
  geom_line(aes(group=bearID, color=bearID), lwd=1, alpha=0.6)+
  geom_jitter(aes(color=bearID, pch=fLMC), size=3, alpha=0.5,width = 0.03, height = 0.01)


```


## Hypothesis #2: Explaining variation in cortisol

This model predicts that there are detectable physiological differences between reproducing females, females newly separated from offspring, and females who failed to reproduce

cortisol=b0+(b1xrepro)+(b2xd15N)+(b3xcondition)|year+bearID

### Data Exploration

#### categorical variables

```{r}
hyp2.dat <- subset(bear.reprodata, repro!="WithMother")

# reproductive status
table(hyp2.dat$repro)
# 146 females with cubs versus 139 solitary females

dotchart(hyp2.dat$cortisol.kitcorrected)
# There are 3 outliers that I need to be aware of 


```


```{r}

 Hyp1<- inla(cortisol.kitcorrected~1+repro+dN15+dC13+condition+year+f(
  bearID, model = "iid"
), data = Hyp1.data
  )

summary(Hyp1)

```

