---
title: "Exploring Bayesian Linear Regression Analysis"
author: "Ashlee Mikkelsen"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

rm(list = ls())
gc()
setwd("~/Rprojects/Diet-and-reproduction-in-rown-bears")

library(ggplot2)
library(dplyr)
library(GGally)
library(viridis)
library(lme4)

P <- palette(viridis(8))
bear.reprodata <- read.csv("bearreprodata.csv")
bear.reprodata$fYEAR <- as.factor(bear.reprodata$fYEAR)


```

I previously determine that my data has some auto-correlation and underlying variance structures that violate the assumptions of linear regression modeling. I am now going to explore using Bayesian methods and MCMC to model my data.

The Zuur book is a bit outdated for Bayesian modeling (just the code examples of course). So rather than follow CH. 23 in Zuur et al. 2009
This .Rmd follows the github tutorial by Drs Hartig and Dorman from APES: http://biometry.github.io/APES//LectureNotes/StatsCafe/Linear_models_jags.html

```{r}




```




