---
title: "PLSR tutorial"
author: "Ashlee Mikkelsen"
date: '2022-10-14'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro
In this file I follow a PLSR tutorial by Bertrand *et al*., 2020 

https://fbertran.github.io/plsRglm/


# Setup

```{r}
rm(list=ls())

setwd("~/Rprojects/Diet-and-reproduction-in-rown-bears/Hypothesis2")
```


##load libraries
```{r}
library(plsRglm)
library(ggplot2)
library(wiqid)
library(plsdof)

```

## set graph theme
```{r}
mytheme <- theme(
    axis.text = element_text(size = 18,face = "bold"),
    axis.title = element_text(size = 20, face = "bold"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey96"),
    panel.background = element_rect(fill = "white"),
    axis.line = element_line(colour = "black",size = 1),
    axis.ticks = element_line(size = 1),
    )
theme_set(mytheme)
```

### create data table
```{r}
HP2.dat <- read.csv("BearReproData_H2Analysis.csv")

Cort.Repro <- HP2.dat[,c(13,3,6,10,11,12,14,15,16,22,23,26,28)]
Cort.Repro <- subset(Cort.Repro, reprocat!="Wmother")
Cort.Repro$conditionsq <- Cort.Repro$condition^2
Cort.Repro$age2<-Cort.Repro$age^2

colSums(is.na(Cort.Repro))

ggplot(data = Cort.Repro, aes(cort))+
  geom_histogram(binwidth = sd(Cort.Repro$cort)*0.5)

```


### standardize cont. vars
```{r}
Zcort <- standardize(Cort.Repro$cort)
Zyr <- standardize(Cort.Repro$year)
Zage <- standardize(Cort.Repro$age)
Zage.sq <- standardize(Cort.Repro$age2)
Zncubs <- standardize(Cort.Repro$cubsfirstcount)
Z13C<-standardize(Cort.Repro$C13.suess)
Z15N <- standardize(Cort.Repro$N15)
Zcond <- standardize(Cort.Repro$condition)
Zcond.sq <- standardize(Cort.Repro$conditionsq)
```


### Creat model DF
```{r}
DF.model <- as.data.frame(cbind(Zyr,Zage,Zage.sq,Zncubs,Z13C,
                                Z15N,Zcond,Zcond.sq,Zcort))

```

# Run Models

## first model

```{r}
cv.modpls<-cv.plsR(Zcort~.,
                   data=DF.model,
                   nt=10,K=6)

res.cv.modpls<-cvtable(summary(cv.modpls))

```

## second model

```{r}
res6<-plsR(Zcort~.,
           data=DF.model,
           nt=6, typeVC="standard", pvals.expli=TRUE)

colSums(res6$pvalstep)

res6$InfCrit

cv.modpls<-cv.plsR(Zcort~.,
                   data=DF.model
                   ,nt=10,K=6,NK=200,
                   random=TRUE,verbose = FALSE)

res.cv.modpls=cvtable(summary(cv.modpls))

plot(res.cv.modpls)

```


# Fot the model with 1 component to obtain regression coefficients

```{r}

res<-plsR(Zcort~.,
          data= DF.model,nt=1,
          pvals.expli=TRUE)
res

res$wwetoile

biplot(res6$tt,res6$pp)

modpls2 <- plsR(Zcort~.,
                data= DF.model,
                6,sparse=TRUE)

modpls3 <- plsR(Zcort ~.,
                data=DF.model,6,
                sparse=TRUE,sparseStop=FALSE)

```

# Bootstrapping

```{r}

MvarRepro.bootYX1=bootpls(res,R=5000,verbose=FALSE)

boxplots.bootpls(MvarRepro.bootYX1,indice=2:9)

temp.ci=confints.bootpls(MvarRepro.bootYX1,indice=2:9)
plots.confints.bootpls(temp.ci,typeIC="BCa",colIC=c("blue","blue","blue","blue"), legendpos ="topright")

plot(MvarRepro.bootYX1,index=2,jack=TRUE)

car::dataEllipse(MvarRepro.bootYX1$t[,2], MvarRepro.bootYX1$t[,3], cex=.3, levels=c(.5, .95, .99), robust=T, xlab="X2", ylab="X3")

```

Using another method of bootstrapping to compare

```{r}
MvarRepro.bootYT1=bootpls(res,typeboot="fmodel_np",R=1000)

boxplots.bootpls(MvarRepro.bootYT1,indices=2:9)

temp.ci=confints.bootpls(MvarRepro.bootYT1,indices=2:9)
plots.confints.bootpls(temp.ci,typeIC="BCa",colIC=c("blue","blue","blue","blue"), legendpos ="topright")

```


```{r}

res2<-plsR(Zcort~.,
           data= DF.model,nt=2)

MvarRepro.bootYT2=bootpls(res2,typeboot="fmodel_np",R=1000)
temp.ci2<-confints.bootpls(MvarRepro.bootYT2,indices=2:9)

ind.BCa.MvarReproYT1 <- (temp.ci[,7]<0&temp.ci[,8]<0)|(temp.ci[,7]>0&temp.ci[,8]>0)

ind.BCa.MvarReproYT2 <- (temp.ci2[,7]<0&temp.ci2[,8]<0)|(temp.ci2[,7]>0&temp.ci2[,8]>0)

(matind=(rbind(YT1=ind.BCa.MvarReproYT1,YT2=ind.BCa.MvarReproYT2)))

res2

C <- res2$Coeffs
C

```


# Interrpreting output
Does it make biological sense

## Find means for each covariate
They should be zero, but...
```{r}
mZyr <- mean(Zyr)
mZage <- mean(Zage)
mZage.sq <- mean(Zage.sq)
mZncubs <- mean(Zncubs)
mZ13C <- mean(Z13C)
mZ15N <- mean(Z15N)
mZcond <-mean(Zcond)
mZcond.sq <- mean(Zcond.sq)

```


## Create new data
```{r}

New.cort <- seq(min(Zcort),max(Zcort), length.out=nrow(Cort.Repro))
New.yr <- seq(min(Zyr),max(Zyr), length.out=nrow(Cort.Repro))
New.age <- seq(min(Zage),max(Zage), length.out=nrow(Cort.Repro))
New.age.sq <-New.age^2
New.ncubs <- seq(min(Zncubs),max(Zncubs), length.out=nrow(Cort.Repro))
New.13C <- seq(min(Z13C),max(Z13C), length.out=nrow(Cort.Repro))
New.15N <- seq(min(Z15N),max(Z15N), length.out=nrow(Cort.Repro))
New.cond <- seq(min(Zcond),max(Zcond), length.out=nrow(Cort.Repro))
New.cond.sq <- New.cond^2

```



The regression model is:

Zcort~b0+b1xZyr+b2xZage+b3xZage.sq+b4xZncubs+b5xZ13C+b6xZ15N+b7xZcond+b8xZcond.sq

To look at the relationship between CORT and any given factor, I need to hold all others at their means

## Specific covariate outputs

### year
```{r}
Cort.Yr <- function(x){
  C[1]+C[2]*x+C[3]*mZage+C[4]*mZage.sq+C[5]*mZncubs+C[6]*mZ13C+C[7]*mZ15N+
    C[8]*mZcond+C[9]*mZcond.sq
}
Y <- Cort.Yr(New.yr)
plot(New.yr,Y)

UnZcort.Y <- (Y*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.Y)

UnZyr<- (New.yr*sd(Cort.Repro$year))+mean(Cort.Repro$year)
print(UnZyr)

Year.DF <- as.data.frame(cbind(Y, New.yr,UnZcort.Y,UnZyr,
                               Cort.Repro$cort, Cort.Repro$year))

ggplot(data = Year.DF, aes(UnZyr,UnZcort.Y))+
  geom_line(lwd=1.2)+
  geom_jitter(aes(V6,V5), size=2, alpha=0.3, width = 0.15)+
  ylab("Hair cortisol")+
  xlab("Year")+
  scale_y_continuous(limits = c(0,35))

ggplot(data = Year.DF, aes(UnZyr,UnZcort.Y))+
  geom_line(lwd=1.2)+
  geom_jitter(aes(V6,V5), size=2, alpha=0.3, width = 0.15)+
  ylab("Hair cortisol")+
  xlab("Year")+
  scale_y_continuous(limits = c(0,15))

```

### Age
```{r}
Cort.age <-  function(x){
  C[1]+
    C[2]*mZyr+
    (C[3]*New.age)+(C[4]*(New.age^2))+
    C[5]*mZncubs+
    C[6]*mZ13C+
    C[7]*mZ15N+
    (C[8]*mZcond)+(C[9]*mZcond.sq)
}

A<-Cort.age(New.age)

plot(New.age,A)

UnZcort.A <- (A*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.A)

UnZage<- (New.age*sd(Cort.Repro$age))+mean(Cort.Repro$age)
print(UnZage)

UnZage.sq <- UnZage^2

Age.DF <- as.data.frame(cbind(A, New.age,
                              UnZcort.A,UnZage,
                              Cort.Repro$cort, Cort.Repro$age))

ggplot(data = Age.DF, aes(New.age,A))+
  geom_line(lwd=1.2)

ggplot(data = Age.DF, aes(UnZage,UnZcort))+
  geom_line()

ggplot(data = Age.DF, aes(UnZage,UnZcort.A))+
  geom_line(lwd=1.2)+
  geom_jitter(aes(V6,V5), size=2, alpha=0.3, width = 0.15)+
  ylab("Hair cortisol")+
  xlab("Bear Age")+
  scale_y_continuous(limits = c(0,35))

ggplot(data = Age.DF, aes(UnZage,UnZcort.A))+
  geom_line(lwd=1.2)+
  geom_jitter(aes(V6,V5), size=2, alpha=0.3, width = 0.15)+
  ylab("Hair cortisol")+
  xlab("Bear Age")+
  scale_y_continuous(limits = c(0,15))


```


### Condition

```{r}

Cort.cond <-function(x){
  C[1]+
    C[2]*mZyr+
    (C[3]*mZage)+(C[4]*mZage^2)+
    C[5]*mZncubs+
    C[6]*mZ13C+
    C[7]*mZ15N+
    C[8]*New.cond+C[9]*(New.cond^2)
}
CC <- Cort.cond(New.cond)
print(CC)
UnZcort.C <- (CC*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.C)

UnZcond<- (New.cond*sd(Cort.Repro$condition))+mean(Cort.Repro$condition)
print(UnZcond)

Cond.DF <- as.data.frame(cbind(CC, New.cond,
                              UnZcort.C,UnZcond,
                              Cort.Repro$cort, Cort.Repro$condition))

ggplot(data = Cond.DF, aes(New.cond,CC))+
  geom_line(lwd=1.2)

ggplot(data = Cond.DF, aes(UnZcond,UnZcort.C))+
  geom_line(lwd=1.2)+
  geom_point(aes(V6,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("Bear Body Condition")+
  scale_y_continuous(limits = c(0,35))

ggplot(data = Cond.DF, aes(UnZcond,UnZcort.C))+
  geom_line(lwd=1.2)+
  geom_point(aes(V6,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("Bear Body Condition")+
  scale_y_continuous(limits = c(0,15))

```


### Litter Size

```{r}

Cort.Ncubs <-function(x){
  C[1]+
    C[2]*mZyr+
    (C[3]*mZage)+(C[4]*mZage^2)+
    C[5]*New.ncubs+
    C[6]*mZ13C+
    C[7]*mZ15N+
    C[8]*mZcond+C[9]*(mZcond^2)
}
LS <- Cort.Ncubs(New.ncubs)
print(LS)
UnZcort.LS <- (LS*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.LS)

UnZncubs<- (New.ncubs*sd(Cort.Repro$cubsfirstcount))+
  mean(Cort.Repro$cubsfirstcount)
print(UnZncubs)

Litter.DF <- as.data.frame(cbind(LS, New.ncubs,
                              UnZcort.LS,UnZncubs,
                              Cort.Repro$cort, Cort.Repro$cubsfirstcount))

ggplot(data = Litter.DF, aes(New.ncubs,LS))+
  geom_line(lwd=1.2)

ggplot(data = Litter.DF, aes(UnZncubs,UnZcort.LS))+
  geom_line(lwd=1.2)+
  geom_jitter(aes(V6,V5), size=2, alpha=0.3,width = 0.15)+
  ylab("Hair cortisol")+
  xlab("Litter Size")+
  scale_y_continuous(limits = c(0,35))

```



### Carbon

```{r}

Cort.d13C <-function(x){
  C[1]+
    C[2]*mZyr+
    (C[3]*mZage)+(C[4]*mZage^2)+
    C[5]*mZncubs+
    C[6]*New.13C+
    C[7]*mZ15N+
    C[8]*mZcond+C[9]*(mZcond^2)
}
dC <- Cort.d13C(New.13C)
print(dC)
UnZcort.dC <- (dC*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.dC)

UnZd13C<- (New.13C*sd(Cort.Repro$C13.suess))+
  mean(Cort.Repro$C13.suess)
print(UnZd13C)

d13C.DF <- as.data.frame(cbind(dC, New.13C,
                              UnZcort.dC,UnZd13C,
                              Cort.Repro$cort,Cort.Repro$C13.suess))

ggplot(data = Litter.DF, aes(New.13C,dC))+
  geom_line(lwd=1.2)

ggplot(data = d13C.DF, aes(UnZd13C,UnZcort.dC))+
  geom_line(lwd=1.2)+
  geom_point(aes(V6,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("d13C")+
  scale_y_continuous(limits = c(0,35))

ggplot(data = d13C.DF, aes(UnZd13C,UnZcort.dC))+
  geom_line(lwd=1.2)+
  geom_point(aes(V6,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("d13C")+
  scale_y_continuous(limits = c(0,15))

```


### Nitrogen

```{r}

Cort.d15N <-function(x){
  C[1]+
    C[2]*mZyr+
    (C[3]*mZage)+(C[4]*mZage^2)+
    C[5]*mZncubs+
    C[6]*mZ13C+
    C[7]*New.15N+
    C[8]*mZcond+C[9]*(mZcond^2)
}
dN <- Cort.d15N(New.15N)
print(dN)
UnZcort.dN <- (dN*sd(Cort.Repro$cort))+mean(Cort.Repro$cort)
print(UnZcort.dN)

UnZd15N<- (New.15N*sd(Cort.Repro$N15))+
  mean(Cort.Repro$N15)
print(UnZd15N)

d15N.DF <- as.data.frame(cbind(dN, New.15N,
                              UnZcort.dN,UnZd15N,
                              Cort.Repro$cort,Cort.Repro$N15))

ggplot(data = d15N.DF, aes(New.15N,dN))+
  geom_line(lwd=1.2)

ggplot(data = d15N.DF, aes(UnZd15N,UnZcort.dN))+
  geom_line(lwd=1.2)+
  geom_point(aes(UnZd15N,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("d15N")+
  scale_y_continuous(limits = c(0,35))

ggplot(data = d15N.DF, aes(UnZd15N,UnZcort.dN))+
  geom_line(lwd=1.2)+
  geom_point(aes(V6,V5), size=2, alpha=0.3)+
  ylab("Hair cortisol")+
  xlab("d15N")+
  scale_y_continuous(limits = c(0,17))

```
