---
title: "MaternalCare Models"
author: "Ashlee Mikkelsen"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Below I begin modelling one aspect of the relationship between physiology and reproduction in Brown bears in a monitored population in south-central Sweden: Variation in maternal care.

The length of maternal care varies between family groups that stick together for 1.5 years and those who stick together for 2.5 years. There are evolutionary advantages to both strategies, and some females with between them and some always use the same strategy throughout their lives. Because brown bears have one of the lowest reproductive output on earth, variation in time between litters is important to population dynamics.

I hypothesized that the balance between energetic demands and resource availability will affect reproductive fitness through the length of maternal care. Length of Maternal care may be affected by cortisol concentrations, d15N, female body condition, litter size, and offspring condition

## Basic setup
Load data and packages, set working directory

```{r}

setwd("~/Rprojects/Diet-and-reproduction-in-rown-bears/Univariate Models-First attempt")

rm(list = ls())

library(rjags)
library(ggplot2)
library(viridis)
library(jagsUI)
library(wiqid)

Bear.ReproData <- read.csv("BearReproData_Analysis.csv")

MatCare <- subset(Bear.ReproData, weaningage!="unknown")
MatCare <- subset(MatCare, statut!="unknown")

```

## Binomial Maternal Care

I tried diving into multinomial modeling and got a bit lost in the weeds. So I am taking a step back and beginning with a basic logistic regression model. To do this, I will drop the litter loss category and only model the probability between length of maternal care of 2.5 years versus 1.5 years.

### Data prep

```{r}
# Subset data to limit results to 2 categories
DFbinom <-subset(MatCare, weaningage!="litterloss") 

##### Covariate setup

# For the categorical variable of parity, Primiparous (P) will be the refernce and multiparous will be the effect. 

# create dummy variables for each yi to determine which parity they belong to
MP <- as.numeric(DFbinom$statut=="M")
Zcort <- standardize(DFbinom$cort)

# create variable to hold the model outcome (length of maternal care). length of 1.5 is more common, so I will let that be the reference and longer maternal care the effect
y <- as.numeric(DFbinom$weaningage=="2.5")

# Dim all have same length
N <- length(P)


```

### Model Specification

```{r}

cat(
  "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0+b1*Zcort[i]+b2*MP[i]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  ## Priors
  p0 ~ dbeta(1,1)
  b0 <- logit(p0)
  b1 ~ dunif(-5,5)
  b2 ~ dunif(-5,5)
  
  } # end model code
  ",
  file = "BinomMCmodel_basic.txt"
)


```



### Fit model, check convergence, and save output

Below I run and fit the JAGS model

```{r}

BinomMC.data <- list(N=N, y=y,Zcort=Zcort,MP=MP)
str(BinomMC.data)

wanted <- c("b0","b1","b2")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

JagsBinomMC <- jags(BinomMC.data,NULL,wanted,
                    model="BinomMCmodel_basic.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 10000,
                    n.burnin = 2000, n.thin = 10, DIC = TRUE,
                    parallel = TRUE)


```

And below I check convergence


```{r}

##### Check convergence
JagsBinomMC_output <- mcmcOutput(JagsBinomMC)

summary(JagsBinomMC_output)

diagPlot(JagsBinomMC_output)

postPlot(JagsBinomMC_output)

```

The model didn't really converge, and we can look at the beta estimates and say something about them, but there isn't a point because this is only an exercise



## Adding Random effect of bearID to model

There are several bears for which we have repeated measures and the number of repeated measures for each bear varies from 1 to 6 repeated measures. This data set contains 138 records for only 67 individual female bears.

```{r}

table(DFbinom$bearID)

bear <- as.factor(MatCare$bearID)
Nbear <- nlevels(bear)
bear <- as.numeric(bear)
sort(bear)


```


### BearID RE model
```{r}
cat(
  "
  model{  # Begin Model code
  # Likelihood
  for(i in 1:N){  ## begin likelihood loop
  
  ## Predictors
  logit(pi2[i]) <- b0[bear[i]]+b1*Zcort[i]+b2*r2[i]
  logit(pi3[i]) <- b0[bear[i]]+b1*Zcort[i]+b2*r2[i]
  
  ## outcome levels 2 and 3 (2.5 and litterloss)
  o2[i]~dbern(pi2[1])
  o3[i]~dbern(pi3[i])
  
  } ## end likelihood loop
  
  ## Priors
  
  for(j in 1:Nbear){  ### begin random intercept loop
  b0[j]~dnorm(mu,tau)
  } ### end random intercept loop
  
  b1~dnorm(0,0.001)
  b2~dnorm(0,0.001)
  
  ## hyperpriors
  pMean ~ dbeta(1,1)
  mu <- logit(pMean)
  sigma ~ dunif(0,5)
  tau <- 1/sigma^2
  } # end model code
  ",
  file = "BasicMCmodel_BearID.txt"
)


```

### Fit model, check convergence, and save output

```{r}

ReBasicMC.data <- list("N"=N,"o2"=o2,"o3"=o3,"Zcort"=Zcort,"r2"=r2,
                       "Nbear"=Nbear,"bear"=bear)
str(ReBasicMC.data)

REwanted <- c("b0","b1","b2", "mu", "sigma")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

ReJagsBasicMC <- jags(ReBasicMC.data,NULL,REwanted,
                    model="BasicMCmodel_BearID.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 5000,
                    n.burnin = 1000, n.thin = 5,
                    parallel = TRUE)


```

check convergence

```{r}

##### Check convergence
ReJagsBasicMC_output <- mcmcOutput(ReJagsBasicMC)

summary(ReJagsBasicMC_output)

diagPlot(ReJagsBasicMC_output)

postPlot(ReJagsBasicMC_output)


```







## Multinomial Maternal Care Model

### Data Prep
Below is my first attempt to write a multinomial logistic regression in JAGS to model one continuous (cort) and one categorical (parity) variable on the probability of a female spending 1.5 years with cubs vs. 2.5 years with cubs vs. lost litters

```{r}

## Step 1. Data setup
### input values. Note below I avoid using the Primiparous (P)category because it is our reference level and the constant absorbs the reference

# here we create dummy variables for each yi to determine which parity they belong to
r1 <- as.numeric(MatCare$statut=="P")
r2 <- as.numeric(MatCare$statut=="M")
Zcort <- standardize(MatCare$cort)

# Here I create variables to hold the model outcome variables (length of maternal care)
o1 <- as.numeric(MatCare$weaningage=="1.5")
o2 <- as.numeric(MatCare$weaningage=="2.5")
o3 <- as.numeric(MatCare$weaningage=="litterloss")

# Dim all have same length
N <- length(r2)


```

### Model Specification

```{r}

cat(
  "
  model{  # Open bracket1
  for(i in 1:N){  ##open bracket2
  ## outcome levels 2 and 3 (2.5 and litterloss)
  o1[i]~dbern(pi1[i])
  o2[i]~dbern(pi2[1])
  o3[i]~dbern(pi3[i])
  
  ## Predictors
  logit(pi1[i]) <- b0+b1*Zcort[i]+b2*r2[i]
  logit(pi2[i]) <- b0+b1*Zcort[i]+b2*r2[i]
  logit(pi3[i]) <- b0+b1*Zcort[i]+b2*r2[i]
  } ## closed bracket 2
  
  ## Priors
  b0~dnorm(0,0.001)
  b1~dnorm(0,0.001)
  b2~dnorm(0,0.001)
  } #closed bracket1
  ",
  file = "BasicMCmodel.txt"
)

cat()

```

### Fit model, check convergence, and save output

Below I run and fit the JAGS model

```{r}

BasicMC.data <- list("r2"=r2,"o1"=o1,"o2"=o2,"o3"=o3,"Zcort"=Zcort,"N"=N)
str(BasicMC.data)

wanted <- c("b0","b1","b2")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

JagsBasicMC <- jags(BasicMC.data,NULL,wanted,
                    model="BasicMCmodel.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 5000,
                    n.burnin = 1000, n.thin = 5, DIC = TRUE,
                    parallel = TRUE)


```

And below I check convergence


```{r}

##### Check convergence
JagsBasicMC_output <- mcmcOutput(JagsBasicMC)

summary(JagsBasicMC_output)

diagPlot(JagsBasicMC_output)

postPlot(JagsBasicMC_output)


```


## Adding Random effect of bearID to model

There are several bears for which we have repeated measures and the number of repeated measures for each bear varies from 1 to 8 repeated measures. This data set contains 152 records for only 67 individual female bears.

```{r}

table(MatCare$bearID)

bear <- as.factor(MatCare$bearID)
print(Nbear)
Nbear <- nlevels(bear)
bear <- as.numeric(bear)
sort(bear)
print(bear)


```


### BearID RE model
```{r}
cat(
  "
  model{  # Begin Model code
  # Likelihood
  for(i in 1:N){  ## begin likelihood loop
  
  ## Predictors
  logit(pi2[i]) <- b0[bear[i]]+b1*Zcort[i]+b2*r2[i]
  logit(pi3[i]) <- b0[bear[i]]+b1*Zcort[i]+b2*r2[i]
  
  ## outcome levels 2 and 3 (2.5 and litterloss)
  o2[i]~dbern(pi2[1])
  o3[i]~dbern(pi3[i])
  
  } ## end likelihood loop
  
  ## Priors
  
  for(j in 1:Nbear){  ### begin random intercept loop
  b0[j]~dnorm(mu,tau)
  } ### end random intercept loop
  
  b1~dnorm(0,0.001)
  b2~dnorm(0,0.001)
  
  ## hyperpriors
  pMean ~ dbeta(1,1)
  mu <- logit(pMean)
  sigma ~ dunif(0,5)
  tau <- 1/sigma^2
  } # end model code
  ",
  file = "BasicMCmodel_BearID.txt"
)


```

### Fit model, check convergence, and save output

```{r}

ReBasicMC.data <- list("N"=N,"o2"=o2,"o3"=o3,"Zcort"=Zcort,"r2"=r2,
                       "Nbear"=Nbear,"bear"=bear)
str(ReBasicMC.data)

REwanted <- c("b0","b1","b2", "mu", "sigma")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

ReJagsBasicMC <- jags(ReBasicMC.data,NULL,REwanted,
                    model="BasicMCmodel_BearID.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 5000,
                    n.burnin = 1000, n.thin = 5,
                    parallel = TRUE)


```

check convergence

```{r}

##### Check convergence
ReJagsBasicMC_output <- mcmcOutput(ReJagsBasicMC)

summary(ReJagsBasicMC_output)

diagPlot(ReJagsBasicMC_output)

postPlot(ReJagsBasicMC_output)


```


