---
title: "MaternalCare Models"
author: "Ashlee Mikkelsen"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction

Below I begin modelling one aspect of the relationship between physiology and reproduction in Brown bears in a monitored population in south-central Sweden: Variation in maternal care.

The length of maternal care varies between family groups that stick together for 1.5 years and those who stick together for 2.5 years. There are evolutionary advantages to both strategies, and some females with between them and some always use the same strategy throughout their lives. Because brown bears have one of the lowest reproductive output on earth, variation in time between litters is important to population dynamics.

I hypothesized that the balance between energetic demands and resource availability will affect reproductive fitness through the length of maternal care. Length of Maternal care may be affected by cortisol concentrations, d15N, female body condition, litter size, and offspring condition

### Basic setup
Load data and packages, set working directory

```{r}

setwd("~/Rprojects/Diet-and-reproduction-in-rown-bears/Univariate Models-First attempt")

rm(list = ls())

library(rjags)
library(ggplot2)
library(viridis)
library(jagsUI)
library(wiqid)
library(GGally)
library(MCMCvis)

Bear.ReproData <- read.csv("BearReproData_Analysis.csv")

MatCare <- subset(Bear.ReproData, reprocat=="Wcubs")
str(MatCare)
colSums(is.na(MatCare))


```

# Step 1: Building a basic Models
### Data prep

```{r}
# Subset data to limit results to 2 categories
DFbinom <-subset(MatCare, weaningage!="litterloss") 

##### Covariate setup

# For the categorical variable of parity, Primiparous (P) will be the refernce and multiparous will be the effect. 

# create dummy variables for each yi to determine which parity they belong to
MP <- as.numeric(DFbinom$statut=="M")
Zcort <- standardize(DFbinom$cort)

# create variable to hold the model outcome (length of maternal care). length of 1.5 is more common, so I will let that be the reference and longer maternal care the effect
y <- as.numeric(DFbinom$weaningage=="2.5")

# Dim all have same length
N <- length(MP)

```

### Model Specification

```{r}

cat(
  "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0+b1*Zcort[i]+b2*MP[i]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  ## Priors
  p0 ~ dbeta(1,1)
  b0 <- logit(p0)
  b1 ~ dunif(-5,5)
  b2 ~ dunif(-5,5)
  
  } # end model code
  ",
  file = "BinomMCmodel_basic.txt"
)

```



### Fit model, check convergence, and save output

Below I run and fit the JAGS model

```{r}

BinomMC.data <- list(N=N, y=y,Zcort=Zcort,MP=MP)
str(BinomMC.data)

wanted <- c("b0","b1","b2")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

JagsBinomMC <- jags(BinomMC.data,NULL,wanted,
                    model="BinomMCmodel_basic.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 10000,
                    n.burnin = 2000, n.thin = 10, DIC = TRUE,
                    parallel = TRUE)

```

And below I check convergence


```{r}

##### Check convergence
JagsBinomMC_output <- mcmcOutput(JagsBinomMC)

summary(JagsBinomMC_output)

diagPlot(JagsBinomMC_output)

postPlot(JagsBinomMC_output)

```

The model didn't really converge, and we can look at the beta estimates and say something about them, but there isn't a point because this is only an exercise



# Step 2: Adding Random effects to model

There are several bears for which we have repeated measures and the number of repeated measures for each bear varies from 1 to 6 repeated measures. This data set contains 139 records for only 60 individual female bears.

### BearID RE model

```{r}

table(DFbinom$bearID)

bear <- as.factor(DFbinom$bearID)
Nbear <- nlevels(bear)
bear <- as.numeric(bear)
sort(bear)

```



```{r}
cat(
  "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0[bear[i]]+b1*Zcort[i]+b2*MP[i]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  ## Priors
  
  for(j in 1:Nbear){  ### begin random intercept loop
  b0[j]~dnorm(mu,tau)
  } ### end random intercept loop
  
  b1 ~ dunif(-5,5)
  b2 ~ dunif(-5,5)
  
  ## hyperpriors
  pMean ~ dbeta(1,1)
  mu <- logit(pMean)
  sigma ~ dunif(0,5)
  tau <- 1/sigma^2
  
  } # end model code
  ",
  file = "ReBinomMCmodel_BearID.txt"
)

```

#### Fit model, check convergence, and save output

```{r}

ReBinomMC.data <- list(N=N, y=y,Zcort=Zcort,MP=MP, Nbear=Nbear, bear=bear)
str(ReBinomMC.data)

wanted <- c("b0","b1","b2", "mu","sigma")

ReJagsBinomMC <- jags(ReBinomMC.data,NULL,wanted,
                    model="ReBinomMCmodel_BearID.txt",
                    n.chains = 4,n.adapt = 2000,n.iter = 10000,
                    n.burnin = 2000, n.thin = 10, DIC = TRUE,
                    parallel = TRUE)

```

check convergence

```{r}

##### Check convergence
ReJagsBinomMC_output <- mcmcOutput(ReJagsBinomMC)

summary(ReJagsBinomMC_output)

diagPlot(ReJagsBinomMC_output)

postPlot(ReJagsBinomMC_output)


```


### Year & Bear ID RE model

#### create year variable
```{r}

table(DFbinom$year)

yr <- as.factor(DFbinom$year)
Nyr <- nlevels(yr)
yr <- as.numeric(yr)

```


#### write model code
```{r}
cat(
  "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0 + b1*Zcort[i]+b2*MP[i] + eps_bear[bear[i]]+
  eps_year[yr[i]]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  ## Priors
  
  for(i in 1:Nbear){  ### begin individual random intercept loop
  eps_bear[i]~dnorm(0,bear_tau)
  } ### end individual random intercept loop
  
  for(i in 1:Nyr){  ### begin individual random intercept loop
  eps_year[i]~dnorm(0,year_tau)
  } ### end individual random intercept loop
  
  b0 ~ dnorm(0, 0.368) #a bit of a better uniform prior for logit space
  b1 ~ dnorm(0, 0.368)
  b2 ~ dnorm(0, 0.368)
  
  ## hyperpriors
  
  ### individual bear RI
  bsigma ~ dunif(0,5)
  bear_tau <- pow(bsigma,-2)
  
  ### year RI
  ysigma ~ dunif(0,5)
  year_tau <- pow(ysigma,-2)
  
  } # end model code
  ",
  file = "ReBinomMCmodel_YR_BearID.txt"
)

```

#### Fit model, check convergence, and save output

```{r}

ReBinom_YrBID.data <- list(N=N, y=y,Zcort=Zcort,MP=MP, Nbear=Nbear,bear=bear,
                       Nyr=Nyr,yr=yr)
str(ReBinom_YrBID.data)

wanted <- c("b0","b1","b2", "bsigma","ysigma")


ReJagsBinomMC <- jags(ReBinom_YrBID.data,NULL,wanted,
                    model="ReBinomMCmodel_YR_BearID.txt",
                    n.chains = 4,n.adapt = 20000,n.iter = 1000000,
                    n.burnin = 20000, n.thin = 50, DIC = TRUE,
                    parallel = TRUE)

```

check convergence

```{r}

##### Check convergence
ReJagsBinomMC_output <- mcmcOutput(ReJagsBinomMC)

summary(ReJagsBinomMC_output)
Pb0<-exp(-2.155)/(1+exp(-2.155))

diagPlot(ReJagsBinomMC_output)

postPlot(ReJagsBinomMC_output)


```

# Step 3: Adding more covariates

I successfully built and ran a logistic model with one random effect, one continuous fixed effect, and one categorical fixed effect. I now need to complete my model. I need to include a random effect of year, because initial data exploration revealed annual variation in diet and cortisol. Annual variation may also contribute to variation in length of maternal care. I also need to include the fixed effects: d15N, d13C, body condition, litter size, and mean offspring condition. In the future, I will also need to include individual diet estimates, but that will be after I finally get MixSIAR to converge.

```{r}
#################
## Rename variables for easier typing
colnames(DFbinom)[11] <- "LitSize"
colnames(DFbinom)[15] <- "LMC"
colnames(DFbinom)[22] <- "d13C"
colnames(DFbinom)[23] <- "d15N"
colnames(DFbinom)[26] <- "MatCond"
colnames(DFbinom)[28] <- "MatCort"
colnames(DFbinom)[29] <- "LitCond"
colnames(DFbinom)[30] <- "LitCort"

```


## Covariate relationships

### Do I expect any interactions?
In preparing the full model, I need to determine if there are any important interactions that I need to model. 

There could be an interaction between body condition and litter size. Females may have too large of a litter, requiring longer maternal care while females in good condition may have the same litter size but seperate sooner. Therefore, I need to include (body condition*littersize).

LMC = b0+ b1 x Cortisol + b2 x N15 + b3 x C13+ b4 x MaternalBC + b5 x LitterSize + b6 x LitterBC + b7 x age + b8 x (MaternalBC x LiiterSize) 

For random effects, I need to include BearID and Year

### Checking Independent variables for colinearity and violations of homogeneity and underlying structure in the data.

#### Colinearity

```{r}
BiNomMC.NumVar <- DFbinom[,c(3,6,11,22,23,26,28,29)]

pairs(BiNomMC.NumVar)

ggpairs(BiNomMC.NumVar)

res <- cor(BiNomMC.NumVar, method = c("pearson", "kendall", "spearman"))
print(res)
corrplot::corrplot(res)

# Maternal cort is highly correlated with year.
ggplot(data = BiNomMC.NumVar, aes(year, MatCort))+
  geom_jitter(width = 0.2)+
  theme_classic()

# Maternal cond is highly correlated with year.
ggplot(data = BiNomMC.NumVar, aes(year, MatCond))+
  geom_jitter(width = 0.2)+
  theme_classic()

# Litter size is correlated with age
ggplot(data = BiNomMC.NumVar, aes(age, LitSize))+
  geom_jitter(width = 0.2)+
  theme_classic()
# Eh, It's mostly driven by 5 and 6 yr old bears. And this is no surprise and not really concerning.


# Litter cond is highly correlated with year. And follows the same pattern as maternal condition
ggplot(data = BiNomMC.NumVar, aes(year, LitCond))+
  geom_jitter(width = 0.2)+
  theme_classic()

# Maternal condition and maternal cort are weak to moderately correlated
# Maternal cond is highly correlated with year.
ggplot(data = BiNomMC.NumVar, aes(MatCort, MatCond))+
  geom_jitter(width = 0.002)+
  theme_classic()

```

#### Cleveland dot plots

```{r}

#######################
#year
dotchart(DFbinom$year,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="year",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )
# There are plenty of years in which we do not have values for 2.5, but since I am not modeling this as a trend, but only a random effect, this should not be a problem. I will need to turn year into a factor though.

DFbinom$fYR <- as.factor(DFbinom$year)

#######################
#age
dotchart(DFbinom$age,
         groups = factor(DFbinom$LMC),
         ylab= "Length Maternal Care",
         xlab="age",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )
ggplot(data = DFbinom, aes(age))+
  geom_histogram(aes(fill=LMC), binwidth = 1)+
  scale_fill_viridis(discrete = TRUE, begin = 0.2, end = 0.8)+
  theme_classic()+
  scale_x_continuous(breaks = seq(4,24,1))

# sampe size decreases with age, which isn't surprising. We only have samples for both lengths of MC for ages 5-18. I would be fine in trimming my data down
DFbinom2 <- subset(DFbinom, age > 4)
DFbinom <- subset(DFbinom2, age < 19)

#######################
# litter size
dotchart(DFbinom$LitSize,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="Litter Size",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )

# Fantastic. We have an uneven distribution of samples within categories, but theyare at least represented within each litter size

#######################
# Carbon
dotchart(DFbinom$d13C,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="d13C",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )

#######################
# Nitrogen
dotchart(DFbinom$d15N,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="d15N",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )

#######################
# Maternal Condition
dotchart(DFbinom$MatCond,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="Maternal Condition",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )

#######################
# Maternal Cortisol
dotchart(DFbinom$MatCort,
         groups = factor(DFbinom$LMC),
         ylab= "LengthMaternalCare",
         xlab="Maternal Cortisol",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )
# I have two of those notorious outliers in this dataset. I need to keep that in mind and should compare the final model results with an analysis without those two outliers

#######################
# Litter Condition
dotchart(DFbinom$LitCond,
         groups = factor(DFbinom$LMC),
         ylab= "Length Maternal Care",
         xlab="Litter Condition",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )
# OK. Something is really going on here. 


rows <- sample(nrow(DFbinom))
RandoSort <- DFbinom[rows,]

dotchart(RandoSort$LitCond,
         groups = factor(RandoSort$LMC),
         ylab= "Length Maternal Care",
         xlab="Litter Condition",
         main="Cleveland Dotplot",
         color = factor(RandoSort$LMC)
         )
# It just so happened that the table was ordered by litter condition.

# Litter Condition
dotchart(DFbinom$LitCort,
         groups = factor(DFbinom$LMC),
         ylab= "Length Maternal Care",
         xlab="Litter Cort",
         main="Cleveland Dotplot",
         color = factor(DFbinom$LMC)
         )

colSums(is.na(DFbinom))

# I am missing too many values for litter cort. I am going to have to drop that covariate.

```


#### Transform all continuous variables


```{r}


Zcort <- (DFbinom$MatCort-mean(DFbinom$MatCort))/(2*sd(DFbinom$MatCort))
Zage <- (DFbinom$age-mean(DFbinom$age))/(2*sd(DFbinom$age))
Zlitsize <-(DFbinom$LitSize-mean(DFbinom$LitSize))/(2*sd(DFbinom$LitSize))
Z13C <- (DFbinom$d13C-mean(DFbinom$d13C))/(2*sd(DFbinom$d13C))
Z15N <- (DFbinom$d15N-mean(DFbinom$d15N))/(2*sd(DFbinom$d15N))
Zmatcond <- (DFbinom$MatCond-mean(DFbinom$MatCond))/(2*sd(DFbinom$MatCond))
Zlitcond <-(DFbinom$LitCond-mean(DFbinom$LitCond))/(2*sd(DFbinom$LitCond))

```


# Specifying the full model

## Maternal BC

LMC = b0+ b1 x Cortisol + b2 x N15 + b3 x C13+ b4 x MaternalBC + b5 x LitterSize + b6 x age  + b8 x (MaternalBC x LiiterSize) +(1|BearID) + (1|year)

### Model structure

```{r}

cat(
 "
  model{  # begin model code
    # - scaling continuous predictors so they have mean 0 and sd = 0.5
  # - using a non-informative Cauchy prior dt(0, pow(2.5,-2), 1) 
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0+
                  bCORT*Zcort[i]+
                  bMP*MP[i]+
                  b15N*Z15N[i]+
                  b13C*Z13C[i]+
                  bLitSize*Zlitsize[i]+
                  bMatBC*Zmatcond[i]+
                  bAge*Zage[i]+
                  bMBCxLS*(Zmatcond[i]*Zlitsize[i])+
                  eps_bear[bear[i]]+
                  eps_year[yr[i]]
                  
 
  ## outcome (LMC = 2.5)
  y[i]~dbin(p[i],1)
  
  } ## end likelihood
  
  ## Priors
  
  for(i in 1:Nbear){  ### begin individual random intercept loop
  eps_bear[i]~dnorm(0,bear_tau)
  } ### end individual random intercept loop
  
  for(i in 1:Nyr){  ### begin individual random intercept loop
  eps_year[i]~dnorm(0,year_tau)
  } ### end individual random intercept loop
  
  b0 ~ dt(0, pow(2.5, -2), 1)
  bCORT ~ dt(0, pow(2.5, -2), 1)
  bMP ~ dt(0, pow(2.5, -2), 1)
  b15N ~ dt(0, pow(2.5, -2), 1)
  b13C ~ dt(0, pow(2.5, -2), 1)
  bLitSize ~ dt(0, pow(2.5, -2), 1)
  bMatBC ~ dt(0, pow(2.5, -2), 1)
  bAge ~ dt(0, pow(2.5, -2), 1)
  bMBCxLS ~ dt(0, pow(2.5, -2), 1)
  
  
  ## hyperpriors
  
  ### individual bear RI
  bsigma ~ dunif(0,5)
  bear_tau <- pow(bsigma,-2)
  
  ### year RI
  ysigma ~ dunif(0,5)
  year_tau <- pow(ysigma,-2)
  
  # GOF test

for(i in 1:N){
    presi[i] <- (y[i] - p[i]) / sqrt(p[i] * (1 - p[i]))
    obs.new[i] ~ dbin(p[i], 1)
    presi.new[i] <- (obs.new[i] - p[i]) / sqrt(p[i] * (1 - p[i]))
    D[i] <- pow(presi[i], 2)
    D.new[i] <- pow(presi.new[i], 2)
  }
  fit <- sum(D[])
  fit.new <- sum(D.new[])

  } # end model code
  ",
  file = "RIBinomMCmodel_MBC.txt"
)

```



### Run model

```{r}
# create dummy variables for each yi to determine which parity they belong to
MP <- as.numeric(DFbinom$statut=="M")
# create variable to hold the model outcome (length of maternal care). length of 1.5 is more common, so I will let that be the reference and longer maternal care the effect
y <- as.numeric(DFbinom$LMC=="2.5")
# Dim all have same length
N <- length(MP)

bear <- as.factor(DFbinom$bearID)
Nbear<-nlevels(bear)
bear <- as.numeric(bear)
print(bear)

yr<-as.factor(DFbinom$year)
yr<-as.numeric(yr)
print(yr)


ReBinomMC.MBCdata <- list(N=N, y=y, Zcort=Zcort, MP=MP,
                           bear=bear,Nbear=Nbear, Z15N=Z15N, Z13C=Z13C,
                           Zmatcond=Zmatcond,Zlitsize=Zlitsize, Zage=Zage,
                          yr=yr, Nyr=Nyr)
str(ReBinomMC.MBCdata)

wanted.MBC <- c("b0","bCORT", "bMP", "b15N", "b13C", "bMatBC", "bLitSize",
                "bAge", "bMBCxLS","p","bsigma","ysigma","fit",
                "fit.new","residual","presi")


ReJagsBinomMC.MBC <- jags(ReBinomMC.MBCdata,NULL,wanted.MBC,
                    model="RIBinomMCmodel_MBC.txt",
                    n.chains = 4,n.adapt = 5000,n.iter = 100000,
                    n.burnin = 5000, n.thin = 20, DIC = TRUE,
                    parallel = TRUE)


```

### Check convergence

```{r}

ReJagsBinomMC.MBC_output <- mcmcOutput(ReJagsBinomMC.MBC)

J <- summary(ReJagsBinomMC.MBC_output)
K <- as.data.frame(J[1:9,])
print(K)

write.csv(K,"LMC_MBCoutput.csv")

diagPlot(ReJagsBinomMC.MBC_output)

postPlot(ReJagsBinomMC.MBC_output)

autocorr.diag(as.mcmc(ReJagsBinomMC.MBC_output))


```

### Goodness of fit testing

```{r}

fitstats <- MCMCpstr(ReJagsBinomMC.MBC, params = c("fit", "fit.new"), 
                     type = "chains") 
T.extreme <- fitstats$fit.new >= fitstats$fit
(p.val <- mean(T.extreme))

```


### Model output

```{r}
MCMCsummary(ReJagsBinomMC.MBC, 
            params = c("b0","bCORT", "bMP", "b15N", "b13C",
                       "bMatBC", "bLitSize","bAge", "bMBCxLS"),
            round = 3)
```


## Litter BC
### Model structure

LMC = b0+ b1 x Cortisol + b2 x N15 + b3 x C13 + b5 x LitterSize + b6 x LitterBC + b7 x age  +(1|BearID) + (1|year)



```{r}

cat(
 "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0+
                  bCORT*Zcort[i]+
                  bMP*MP[i]+
                  b15N*Z15N[i]+
                  b13C*Z13C[i]+
                  bLitSize*Zlitsize[i]+
                  bLitBC*Zlitcond[i]+
                  bAge*Zage[i]+
                  eps_bear[bear[i]]+
                  eps_year[yr[i]]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  for(i in 1:Nbear){  ### begin individual random intercept loop
  eps_bear[i]~dnorm(0,bear_tau)
  } ### end individual random intercept loop
  
  for(i in 1:Nyr){  ### begin individual random intercept loop
  eps_year[i]~dnorm(0,year_tau)
  } ### end individual random intercept loop
  
  b0 ~ dt(0, pow(2.5, -2), 1)
  bCORT ~ dt(0, pow(2.5, -2), 1)
  bMP ~ dt(0, pow(2.5, -2), 1)
  b15N ~ dt(0, pow(2.5, -2), 1)
  b13C ~ dt(0, pow(2.5, -2), 1)
  bLitSize ~ dt(0, pow(2.5, -2), 1)
  bLitBC ~ dt(0, pow(2.5, -2), 1)
  bAge ~ dt(0, pow(2.5, -2), 1)
  
  
  ## hyperpriors
  
  ### individual bear RI
  bsigma ~ dunif(0,5)
  bear_tau <- pow(bsigma,-2)
  
  ### year RI
  ysigma ~ dunif(0,5)
  year_tau <- pow(ysigma,-2)

# GOF test

for(i in 1:N){
    presi[i] <- (y[i] - p[i]) / sqrt(p[i] * (1 - p[i]))
    obs.new[i] ~ dbin(p[i], 1)
    presi.new[i] <- (obs.new[i] - p[i]) / sqrt(p[i] * (1 - p[i]))
    D[i] <- pow(presi[i], 2)
    D.new[i] <- pow(presi.new[i], 2)
  }
  fit <- sum(D[])
  fit.new <- sum(D.new[])
  
  } # end model code
  ",
  file = "ReBinomMCmodel_LBC.txt"
)

```


#### Run model

```{r}
# create dummy variables for each yi to determine which parity they belong to
MP <- as.numeric(DFbinom$statut=="M")
# create variable to hold the model outcome (length of maternal care). length of 1.5 is more common, so I will let that be the reference and longer maternal care the effect
y <- as.numeric(DFbinom$LMC=="2.5")
# Dim all have same length
N <- length(MP)

bear <- as.factor(DFbinom$bearID)
Nbear <- nlevels(bear)
bear <- as.numeric(bear)
sort(bear)


ReBinomMC.LBCdata <- list(N=N, y=y,Zcort=Zcort,MP=MP, Nbear=Nbear,
                          bear=bear,Z15N=Z15N, Z13C=Z13C, Zlitsize=Zlitsize,
                          Zlitcond=Zlitcond, Zage=Zage, Nyr=Nyr, yr=yr)
str(ReBinomMC.LBCdata)

wanted.LBC <- c("bCORT", "bMP", "b15N", "b13C", "bLitSize",
            "bLitBC","bAge","bsigma","ysigma","b0","p","fit",
            "fit.new","residual","presi")


ReJagsBinomMC.LBC <- jags(ReBinomMC.LBCdata,NULL,wanted.LBC,
                    model="ReBinomMCmodel_LBC.txt",
                    n.chains = 4,n.adapt = 5000,n.iter = 100000,
                    n.burnin = 5000, n.thin = 20, DIC = TRUE,
                    parallel = TRUE)


```

#### Check convergence

```{r}

ReJagsBinomMC.LBC_output <- mcmcOutput(ReJagsBinomMC.LBC)

J <- summary(ReJagsBinomMC.LBC_output)
K <- as.data.frame(J[1:9,])
print(K)

write.csv(K,"LMC_LBCoutput.csv")

diagPlot(ReJagsBinomMC.LBC_output)

postPlot(ReJagsBinomMC.LBC_output)

autocorr.diag(as.mcmc(ReJagsBinomMC.LBC_output))


```


### Goodness of fit testing

```{r}

fitstats <- MCMCpstr(ReJagsBinomMC.LBC, params = c("fit", "fit.new"), 
                     type = "chains") 
T.extreme <- fitstats$fit.new >= fitstats$fit
(p.val <- mean(T.extreme))

```




# Specifying the reduced model

After running the full model and looking at the output, the only parameters with moderate or greater support was the effect of d15N, maternal body condition, and bear age. I chose to retain the variables and run a reduced model whose output would be less messy:

LMC = b0+ b1 x N15 + b2 x MaternalBC + b3 x age +(1|BearID)

```{r}

cat(
 "
  model{  # begin model code
  
  for(i in 1:N){  ##begin likelihood
  
  ## Predictors
  logit(p[i]) <- b0[bear[i]]+
                  b15N*Z15N[i]+
                  bMatBC*Zmatcond[i]+
                  bAge*Zage[i]
 
  ## outcome (LMC = 2.5)
  y[i]~dbern(p[i])
  
  } ## end likelihood
  
  ## Priors
  
  for(j in 1:Nbear){  ### begin random intercept loop
    b0[j]~dnorm(mu,tau)
  } ### end random intercept loop
  
  
  b15N ~ dunif(-5,5)
  bMatBC ~ dunif(-5,5)
  bAge ~ dunif(-5,5)

  
  ## hyperpriors
  
  pMean ~ dbeta(1,1)
  mu <- logit(pMean)
  sigma ~ dunif(0, 5)
  tau <- 1/sigma^2
  
  
  } # end model code
  ",
  file = "ReBinomMCmodel_Reduced.txt"
)

```


### Run and check model

```{r}
# create dummy variables for each yi to determine which parity they belong to
MP <- as.numeric(DFbinom$statut=="M")
# create variable to hold the model outcome (length of maternal care). length of 1.5 is more common, so I will let that be the reference and longer maternal care the effect
y <- as.numeric(DFbinom$LMC=="2.5")
# Dim all have same length
N <- length(MP)


ReBinomMC.Reduceddata <- list(N=N, y=y,Nbear=Nbear,bear=bear,Z15N=Z15N, 
                              Zmatcond=Zmatcond, Zage=Zage)
str(ReBinomMC.Reduceddata)

wanted.Reduced <- c("b15N", "bMatBC", "bAge", "pMean", "mu", "sigma", "b0")

inits <- function(){
  list(
  b0=rnorm(1,-2,2),
  b1=rnorm(0,-2,2),
  b2=rnorm(1,-2,2)
  )
}

ReJagsBinomMC.Reduced <- jags(ReBinomMC.Reduceddata,NULL,wanted.Reduced,
                    model="ReBinomMCmodel_Reduced.txt",
                    n.chains = 4,n.adapt = 50000,n.iter = 1000000,
                    n.burnin = 50000, n.thin = 50, DIC = TRUE,
                    parallel = TRUE)


```

### Check convergence

```{r}

ReJagsBinomMC.Reduced_output <- mcmcOutput(ReJagsBinomMC.Reduced)

J2 <- summary(ReJagsBinomMC.Reduced_output)
K2 <- as.data.frame(J2[1:9,])
print(K2)

write.csv(K2,"LMC_JagsReducedModelOutput.csv")

diagPlot(ReJagsBinomMC.Reduced_output)

postPlot(ReJagsBinomMC.Reduced_output)

autocorr.diag(as.mcmc(ReJagsBinomMC.Reduced_output))



```


### Interpreting model parameters

```{r}

max(Z15N)
min(Z15N)
NewZ15N <- seq(-2.9,2.9,0.0585)
MZ15N <- rep(mean(Z15N),100)

min(Zmatcond)
max(Zmatcond)
NewZMatCond <- seq(-2.7,2.2, 0.0493)
MZMBC <- rep(mean(Zmatcond),100)

min(Zage)
max(Zage)
NewZage <- seq(-1.3,2.2, 0.0352)
MZAGE <- rep(mean(Zage),100)


LMCage <- vector(length = 100)
LMCmbc <- vector(length = 100)
LMCn <- vector(length = 100)

for (i in 1:100) {
  LMCage[i]=(exp(0.186+
               (0.36*MZ15N[i])+
               (-0.568*MZMBC[i])+
               (1.009*NewZage[i])))/(1+(exp(0.186+
               (0.36*MZ15N[i])+
               (-0.568*MZMBC[i])+
               (1.009*NewZage[i]))))
  
}

for (i in 1:100) {
  LMCmbc[i]=(exp(0.186+
               (0.36*MZ15N[i])+
               (-0.568*NewZMatCond[i])+
               (1.009*MZAGE[i])))/(1+(exp(0.186+
               (0.36*MZ15N[i])+
               (-0.568*NewZMatCond[i])+
               (1.009*MZAGE[i]))))
  
}


for (i in 1:100) {
  LMCn[i]=(exp(0.186+
               (0.36*NewZ15N[i])+
               (-0.568*MZMBC[i])+
               (1.009*MZAGE[i])))/(1+(exp(0.186+
               (0.36*NewZ15N[i])+
               (-0.568*MZMBC[i])+
               (1.009*MZAGE[i]))))
  
}


New.Dat <- cbind(NewZ15N,NewZMatCond,NewZage, LMCmbc, LMCage,LMCn)
New.Dat <- as.data.frame(New.Dat)
Mage <- mean(DFbinom$age)
SDage <- sd(DFbinom$age)
New.Dat$age <- ((New.Dat$NewZage*SDage)+Mage)
Mmbc <- mean(DFbinom$MatCond)
SDmbc <- sd(DFbinom$MatCond)
New.Dat$MBC <- ((New.Dat$NewZMatCond*SDmbc)+Mmbc)
M15n <- mean(DFbinom$d15N)
SDn <- sd(DFbinom$d15N)
New.Dat$d15N <- ((New.Dat$NewZ15N*SDn)+M15n)

mytheme <- theme(
    axis.text = element_text(size = 18,face = "bold"),
    axis.title = element_text(size = 20, face = "bold"),
    panel.grid.major = element_line(color = "grey92"),
    panel.grid.minor = element_line(color = "grey96"),
    panel.background = element_rect(fill = "white"),
    axis.line = element_line(colour = "black",size = 1),
    axis.ticks = element_line(size = 1),
    )



ggplot(data = New.Dat, aes(NewZage,LMCage))+
  geom_line()+
  geom_point()+
  theme_classic()+
  mytheme

ggplot(data = New.Dat, aes(age,LMCage))+
  geom_line(lwd=1.2)+
  theme_classic()+
  scale_y_continuous(breaks = seq(0,1,0.25),
                     limits =c(0,1))+
  scale_x_continuous(expand = c(0,0),
                     breaks = seq(6,18,2))+
  ylab("Probability Of Maternal Care 2.5 years")+
  xlab("Female Age")+
  mytheme


ggplot(data = New.Dat, aes(NewZ15N,LMCn))+
  geom_line()+
  geom_point()+
  theme_classic()

ggplot(data = New.Dat, aes(d15N,LMCn))+
  geom_line(lwd=1.2)+
  theme_classic()+
  scale_y_continuous(breaks = seq(0,1,0.25),
                     limits =c(0,1))+
  scale_x_continuous(expand = c(0,0),
                     breaks = seq(3,6,0.5))+
  ylab("Probability Of Maternal Care 2.5 years")+
  xlab("Female d15N ")+
  mytheme

ggplot(data = New.Dat, aes(NewZMatCond,LMCmbc))+
  geom_line()+
  geom_point()+
  theme_classic()

ggplot(data = New.Dat, aes(MBC,LMCmbc))+
  geom_line(lwd=1.2)+
  theme_classic()+
  scale_y_continuous(breaks = seq(0,1,0.25),
                     limits =c(0,1))+
  scale_x_continuous(expand = c(0,0),
                     breaks = seq(0,2,0.2))+
  ylab("Probability Of Maternal Care 2.5 years")+
  xlab("Female Body Condition")+
  mytheme

```







